---
title: "Central Limit Theorem"
author: "John Blischak"
date: "2019-07-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Reproducing the demonstration of the Central Limit Theorem in
_Master 'Metrics_ (p. 39-42).

Consider a Bernoulli random variable with probability of success of $p = 0.8$.
The sampling distribution of the mean of this distribution approximates a normal
distribution, especially with increasing sample sizes. This is an example of the
Central Limit Theorem.

```{r functions}
# n - number of draws from a Bernoulli random variable with p = 0.8
randvar <- function(n) rbinom(n, size = 1, prob = 0.8)

# Calculate the t-statistic
tstat <- function(x) (mean(x) - 0.8) / (sd(x) / sqrt(length(x)))

# Visualize distribution compared to standard normal
viz <- function(x, ...) {
  hist(x, freq = FALSE, xlab = "t-statistic", ylab = "Fraction",
       xlim = c(-4, 4), ...)
  x <- seq(-4, 4, by = 0.25)
  y <- dnorm(x)
  lines(x, y, col = "red", lty = "dashed")
}
```

Sample size of 10

```{r n10}
draws <- replicate(500000, randvar(10))
tstats <- apply(draws, 2, tstat)
viz(tstats, main = "n = 10")
```

Sample size of 40

```{r n40}
draws <- replicate(500000, randvar(40))
tstats <- apply(draws, 2, tstat)
viz(tstats, main = "n = 40")
```

Sample size of 100

```{r n100}
draws <- replicate(500000, randvar(100))
tstats <- apply(draws, 2, tstat)
viz(tstats, main = "n = 100")
```
