---
title: "Properties of the summation operator"
author: "John Blischak"
date: "2019-05-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Exploring some of the properties of the summation operator described in Appendix
A of Introductory Econometrics by Jeffrey Woolridge.

## Equation A.7

> The sum of squared deviations is the sum of the squared $x_i$ minus $n$ times
the square of $\bar{x}:$

$$ \sum_{i=1}^{n} (x_i - \bar{x})^2 = \sum_{i=1}^{n} x_i^2 - n(\bar{x})^2 $$

First, confirming this computationally with an example.

```{r}
n <- 10
x <- rpois(n, lambda = 5)
xbar <- mean(x)
```

```{r}
sum((x - xbar)^2)
sum(x^2) - n * xbar^2
```

Second, visually exploring this relationship.

```{r plot-a7}
plot(x)
abline(h = xbar, col = "red", lty = "dashed")
for (i in 1:n) lines(x = c(i, i), y = c(x[i], xbar), col = "blue")
```

Third, the algebraic proof from the book:

$$
\begin{align*}
\sum_{i=1}^{n} (x_i - \bar{x})^2 \\
= \sum_{i=1}^{n} (x_i^2 - 2x_i\bar{x} + \bar{x}^2) \\
= \sum_{i=1}^{n} x_i^2 - 2\bar{x}\sum_{i=1}^{n} x_i + n(\bar{x})^2 \\
= \sum_{i=1}^{n} x_i^2 - 2n(\bar{x})^2 + n(\bar{x})^2 \\
= \sum_{i=1}^{n} x_i^2 - n(\bar{x})^2 
\end{align*}
$$ 

The key step for me is the substitution of $n\bar{x}$ for $\sum_{i=1}^{n} x_i$. This relationship is obtained by rearranging the equation for the mean:

$$ \bar{x} = (1/n) \sum_{i=1}^{n} x_i$$

## Equation A.8

$$
\begin{align*}
\sum_{i=1}^{n} (x_i - \bar{x}) (y_i - \bar{y}) \\
= \sum_{i=1}^{n} x_i (y_i - \bar{y}) \\
= \sum_{i=1}^{n} (x_i - \bar{x}) y_i \\
= \sum_{i=1}^{n} x_i y_i - n(\bar{x} \cdot \bar{y})
\end{align*}
$$

First, confirming this computationally with an example.

```{r}
n <- 10
x <- rpois(n, lambda = 5)
y <- rpois(n, lambda = 5)
xbar <- mean(x)
ybar <- mean(y)
```

```{r}
sum((x - xbar) * (y - ybar))
sum(x * (y - ybar))
sum((x - xbar) * y)
sum(x * y) - n * (xbar * ybar)
```

Second, visually exploring this relationship.

```{r plot-a8}
op <- par(mfrow = c(1, 2))
plot(x, y)
abline(h = xbar, col = "red", lty = "dashed")
for (i in 1:n) lines(x = c(x[i], x[i]), y = c(y[i], xbar), col = "blue")
plot(x, y)
abline(v = ybar, col = "red", lty = "dashed")
for (i in 1:n) lines(x = c(x[i], ybar), y = c(y[i], y[i]), col = "purple")
par(op)
```
